---
title: "MR tutorial II"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    collapsed: true
    theme: united
  author: "Joëlle Pasman, Gudmundur Einarsson"
  date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}

# Essential

library(tidyverse)
library(data.table)
library(TwoSampleMR)
library(ggforestplot)

# Helpful (some non-essential code may not work without these)
library(simex)
library(knitr)

# Working directory is root of 
knitr::opts_knit$set(root.dir = "../notebooks")
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, error=FALSE)

```

For your setup, put all the GWAS summary statistics in a sub folder called sumstats in your working directory. In the same directory, create a folder called "results". 

# Part 1: The basics

## Data preparation steps

First, we create the instrument, using a threshold of p<5e-08 and an R2<0.01. Note that the chunk below requires internet access to run. In this example, we use BMI as our exposure. You can replace this with any other trait. Note that if you use your own summary statistics with different column names, you need to modify the code.

```{r Create instruments}

# Extract hits 

BMI <- fread("../sumstats/bmi.txt")

BMIhits <- BMI %>% 
  filter(P<5e-08) %>%
  mutate(Phenotype="BMI", effect_allele=A1, other_allele=A2, eaf=EAF, beta=B, se=SE, pval=P, samplesize=N) 

fwrite(BMIhits, "../sumstats/BMI_hits.txt", sep=" ", quote=F, na=NA)

# Reading in the exposure file & derive independent instruments
# Note that this step requires internet access (function uses online reference data)

BMI_exp_dat <- read_exposure_data("../sumstats/BMI_hits.txt")
BMI_exp_dat <- clump_data(BMI_exp_dat, clump_r2 = 0.01)

BMI_exp_dat %>% 
  summarise(IV_SNPs=n())

```


Next, we prepare and format the summary statistics that we'll use as outcomes. We will check if the instrument SNPs are all present in the outcome data files and harmonize the data frames.


```{r Harmonize exposure-outcome data}

# For each instrument - outcome combination, we look up the instrument SNPs in the outcome data.

files <- c("../sumstats/scz.txt",
           "../sumstats/mdd.txt",
           "../sumstats/healthspan.txt")

outcomedata <- list()

for (f in 1:length(files)) {
              outcomedata[[f]] <- read_outcome_data(snps = BMI_exp_dat$SNP,
              filename = files[f],
              sep = ",",
              snp_col = "SNP",
              beta_col = "B",
              se_col = "SE",
              effect_allele_col = "A1",
              other_allele_col = "A2",
              eaf_col = "EAF",
              pval_col = "P", 
              ncase="N_CAS",
              ncontrol="N_CON",
              samplesize_col="N", 
              phenotype_col="Outcome")
}

####
# Harmonize

exposure_outcome <- list()
for (i in 1:length(outcomedata)) {
  exposure_outcome[[i]] <- harmonise_data(
    exposure_dat = BMI_exp_dat, 
    outcome_dat = outcomedata[[i]]
  )
}

# Keep only filtered SNPs
for (i in 1:length(outcomedata)) {
  exposure_outcome[[i]] <- subset(exposure_outcome[[i]], mr_keep==T)
}

# write harmonized files to tables for information on SNPs
for (i in 1:length(outcomedata)) {
  fwrite(outcomedata[[i]], paste0("../sumstats/BMI_to_", outcomedata[[i]]$outcome[[1]], "_harmonized.txt"))
}

```

## Main analyses

We run MR with and without Steiger-filtering (removing SNPs that explain more variance in the outcome than in the exposure, i.e., have horizontal pleiotropy). Note that the MR-Egger results should only be interpreted if I-squared (see below) is above .9, otherwise SIMEX correction is required.

```{r Main analyses, results='hide'}

mr_res <- exposure_outcome %>%
  map(mr)

# IVW results plot before Steiger filtering

MRres_forplot <- rbindlist(mr_res) %>% 
  filter(method=="Inverse variance weighted")

MRres_plot <- forestplot(MRres_forplot, 
           title="MR results (IVW) before Steiger filtering",
           name=outcome, 
           estimate = b, 
           se=se, 
           colour = exposure,
           xlab="Beta",
           ylab="Outcome"
           )


####
# Steiger filter all analyses

# Need some extra columns to determine R2 for binary traits. If you have a binary exposure, fill in the number of cases, controls, and prevalence, and use log odds units

for (i in 1:length(exposure_outcome)) {
  exposure_outcome[[i]] <- exposure_outcome[[i]] %>% 
    mutate(units.exposure="SD") %>%
    # mutate(units.exposure="log_odds") %>%
    # mutate(ncase.exposure=xxx) %>% 
    # mutate(ncontrol.exposure=xxx) %>% 
    # mutate(prevalence.exposure=xxx) %>% 
    mutate(units.outcome=ifelse(!is.na(ncase.outcome), "log_odds", "SD")) %>% # this step uses the presence of an N cases column to determine if the outcome trait is binary
    mutate(prevalence.outcome=ifelse(!is.na(ncase.outcome), ncase.outcome/samplesize.outcome, NA))
}

# Determine which SNPs explain more variance in the outcome
exposure_outcome <- exposure_outcome %>% 
  map(steiger_filtering)

# Re-run while filtering out SNPs that are in the wrong direction and explain more variance in the outcome than in the exposure (dir=TRUE). Note you can also filter based on p-value but this is often a bit too stringent.

mr_res_st <- list()
for (i in 1:length(exposure_outcome)) {
  mr_res_st[[i]] <- mr(subset(exposure_outcome[[i]], exposure_outcome[[i]]$steiger_dir==TRUE))
}

# Scatter plots after Steiger filtering (subset list of results based on length; if length is 0 there were no snps left after filtering)
# mr_res_st <- keep(mr_res_st, ~ nrow(.) >0)
# exposure_outcome_st <- keep(exposure_outcome, ~ any(.$steiger_dir))

dir.create("../results")
for (i in 1:length(mr_res_st)) {
  res1 <- mr_res_st[[i]]
  dat1 <- exposure_outcome[[i]]
  name <- paste0("scatter_",res1[1,4],"_",res1[1,3],"_steiger")
  plot_path_svg <-  paste0("../results/", name,".svg")
  plot_path_png <-  paste0("../results/", name,".png")
  p_save <- mr_scatter_plot(res1,dat1)[[1]] + 
    theme_bw() + theme(legend.position = "top")
  ggsave(plot = p_save, filename = plot_path_svg)
  ggsave(plot = p_save, filename = plot_path_png)
}

# IVW results plots after Steiger filtering

MRst_forplot <- rbindlist(mr_res_st) %>% 
  filter(method=="Inverse variance weighted")

MRst_plot <- forestplot(MRst_forplot, 
           title="MR results (IVW) after Steiger filtering",
           name=outcome, 
           estimate = b, 
           se=se, 
           colour = exposure,
           xlab="Beta",
           ylab="Outcome"
           )

```

This is an example of the Scatter plot of the effect of BMI on healthspan (which is reverse-coded, so should be interpreted as morbidity/ mortality). The effect is nicely consistent across the different methods, though it seems largely driven by a small number of SNPs. 

![](../results/scatter_BMI_healthspan_steiger.svg)

Results before Steiger filtering:

```{r res}

restable <- rbindlist(mr_res) %>% 
  select(!c(id.exposure, id.outcome))

kable(restable)

MRres_plot

```


Results after Steiger filtering:

```{r res Steiger}


restable <- rbindlist(mr_res_st) %>% 
  select(!c(id.exposure, id.outcome))

kable(restable)
MRst_plot

```

# Part 2: Sensitivity analyses

Below are a number of extra sensitivity analyses to follow up and assess the robustness of the results. Opinions are divided on which are the most important/ crucial. For general purposes one can suffice with the methods included in the TwoSampleMR package, with the addition of Isq and SIMEX correction and maybe MR-PRESSO (part 3). Code for these standard follow-ups is provided below.

## Pleiotropy

First, using MR Egger, we test if there is pleiotropy, for the Steiger corrected results. Steiger filters out specific pleiotropic SNPs (that have a relationship with the outcome), but other sources of pleiotropy may be detected by MR Egger.

```{r Pleiotropy}

# MR-Egger intercept after Steiger filtering.

MEintercept <- list()
for (i in 1:length(exposure_outcome)) {
  res <- data.frame(mr_pleiotropy_test(exposure_outcome[[i]]))
  MEintercept[[i]] <- res
}

MEintercept <- rbindlist(MEintercept)

kable(MEintercept %>% 
  select(exposure, outcome, egger_intercept, se, pval))

```




## R-squared

In the Steiger-filtering step, the R-squared was derived to assess the variance explained by each independent SNP in the exposure. We sum the R-squareds of each instrument to get a sense of instrument strength. Note that due to harmonizing we have different numbers of IVs in each of the analyses, which is why the Rsq for the same exposure varies per outcome. 

```{r R-squared}

# Extract list of R-squared for each dataframe
r2 <- list()
for (i in 1:length(exposure_outcome)) {
  r2[[i]] <- tibble(exposure_outcome[[i]]$outcome[[1]], sum(subset(exposure_outcome[[i]], steiger_dir==T)$rsq.exposure))
}

res_r2 <- (rbindlist(r2))
colnames(res_r2) <- c("outcome", "Rsq")
kable(res_r2)

```

## I-squared and F-statistic

In this code chunk, I-squared is computed to assess if the MR Egger can be interpreted, or needs to be SIMEX-corrected. For each analysis, the heighest (unweighted or weighted) Isq should be interpreted. If it's below .9 but above .6 a SIMEX correction should be performed. If it's below .6 it should not be reported at all. If it's above .9 no correction is needed.

In the same run, we assess instrument strength by estimating the F-statistic. An F above 10 is considered acceptable instrument strength. 

```{r Isq and F statistic}


# This script and the MR-SIMEX function further below are based on Bowden et al.(2016) 

# Isq function

Isq <- function(y,s){
  k          = length(y)
  w          = 1/s^2; sum.w  = sum(w)
  mu.hat     = sum(y*w)/sum.w  
  Q          = sum(w*(y-mu.hat)^2)
  Isq        = (Q - (k-1))/Q
  Isq        = max(0,Isq)
  return(Isq)
}


# Function to paste Isq and F to the data frames

appendIsq <- function(x) {
  x %>% 
      mutate(BetaXG = beta.exposure, seBetaXG = se.exposure, seBetaYG = se.outcome) %>%
      mutate(BXG = abs(BetaXG)) %>%
      mutate(Isq_unweighted=Isq(BXG,seBetaXG)) %>% 
      mutate(Isq_weighted=Isq((BXG/seBetaYG),(seBetaXG/seBetaYG))) %>% 
      mutate(Isq_max=pmax(Isq_unweighted, Isq_weighted)) %>% 
      mutate(Isq_max_weighted=ifelse(Isq_weighted>Isq_unweighted, "weighted", "unweighted")) %>% 
      mutate(Fstatistic = BXG^2/seBetaXG^2) %>% 
      mutate(mFstatistic = mean(Fstatistic))
}

# Apply the function

exposure_outcome_f <- list()
for (i in 1:length(exposure_outcome)) {
  exposure_outcome_f[[i]] <- appendIsq(exposure_outcome[[i]])
}

# Before we continue, filter out the SNPs with the wrong effect direction according to the Steiger analysis

for (i in 1:length(exposure_outcome_f)) {
  exposure_outcome_f[[i]] <- subset(exposure_outcome_f[[i]], steiger_dir==TRUE)
}

# Extract max Isq and F values per outcome, and make columns indicating if F is sufficient and if SIMEX should be applied 

isqs <- list()
for (i in 1:length(exposure_outcome_f)) {
  isqs[[i]] <- exposure_outcome_f[[i]] %>%
      group_by(exposure, outcome) %>% 
      summarise(Isq_max=max(Isq_max), mean_Fstatistic=max(mFstatistic), Isq_weight=first(Isq_max_weighted)) %>% 
      mutate(SIMEX=ifelse(Isq_max>0.6 & Isq_max<0.9, "SIMEX_egger", NA)) %>% 
      mutate(SIMEX=ifelse(Isq_max>0.9, "Standard_egger", SIMEX)) %>%
      mutate(SIMEX=ifelse(Isq_max<0.6, "No_egger", SIMEX)) %>% 
      mutate(Instrument_strength=ifelse(mean_Fstatistic<10, "weak", "sufficient"))
}

isqs <- rbindlist(isqs)

kable(isqs)

saveRDS(exposure_outcome_f, "exposure_outcome_f.rds")

```

In this example, instrument strength is sufficient. For all exposure-outcome combinations, the unweighted I2 is the highest, and MR Egger SIMEX-correction is necessary.

## SIMEX-correction for MR-Egger

The SIMEX correction takes long to run when you run 10,000 iterations (which is recommended) so it is advisable to run this on a cluster. In this example, we run using B=1000.

```{r SIMEX}

exposure_outcome_f <- readRDS("exposure_outcome_f.rds")

# The simex correction function

simexfunction <- function(x) {
  
  Isq_max <- x$Isq_max[1]
  
  if (Isq_max < 0.6 | Isq_max > 0.9) {return(NULL)}
  
  if (x$Isq_max_weighted[[1]] == "weighted") {
    
    x_trans <- x %>% mutate(y = beta.outcome * sign(beta.exposure), exposure = abs(beta.exposure))
    fit <- lm(y ~ exposure, weights=1/se.outcome^2, data = x_trans, x = TRUE, y = TRUE)
    
  }
  
  if (x$Isq_max_weighted[[1]] == "unweighted") {
    
    x_trans <- x %>% mutate(y = beta.outcome, exposure = abs(beta.exposure))
    fit <- lm(y ~ exposure, data = x_trans, x = TRUE, y = TRUE)
    
  }
  
  res <- summary(simex(fit,
                       B = 1000,
                       measurement.error = x$se.exposure,
                       SIMEXvariable = "exposure",
                       fitting.method = "quad",
                       asymptotic = "FALSE"))
  
  res$coefficients
  
}

# Apply the function  
simex_res <- exposure_outcome_f %>%
  map(simexfunction)

outcome_name_vector <- vector()

for (i in 1:length(exposure_outcome_f)) {
  outcome_name_vector[[i]] <- exposure_outcome[[i]]$outcome[[1]]
}

names(simex_res) <- outcome_name_vector

# Results in nexted lists, extract the results and wrangle to dataframe

unnested <- compact(simex_res)
for (i in 1:length(unnested)) {
  unnested[[i]] <- data.frame(unnested[[i]]$jackknife)
  unnested[[i]] <- unnested[[i]] %>% slice(2)
  colnames(unnested[[i]]) <- c("simex_estimate", "simex_SE", "simex_t-value", "simex_p-value")
}

unnested <- rbindlist(unnested, use.names = T, id=T)
unnested <- unnested %>% rename(outcome=.id)

# Add to the dataframe containing the results and isqs
simex_results <- full_join(isqs, unnested)


```
These are the SIMEX-corrected MR-Egger estimates.

``` {r res simex}

kable(simex_results)

```


## Heterogeneity 

The code below assesses heterogeneity using Cochran's Q, leave-one-out analysis, and single-snp analysis. LOO-plots show the consistency of the effects when a single SNP is left out of the instrument set. Single-SNP funnel plots show the effect of each single SNP to visualize if there's  asymmetry, that is: if effects from weaker variants tend to go in one direction. Q is a formal test of such SNP effect outliers. If Q is much larger than the number of instrument SNPs it becomes significant. 

```{r Heterogeneity analyses, echo=F, results='hide'}

# Here, I use the original (unfiltered) data, subsetting them for a correct Steiger-direction, that can be changed if desired.

# Leave-one-out analysis

loo <- list()
for (i in 1:length(exposure_outcome)) {
  loores <- data.frame(mr_leaveoneout(subset(exposure_outcome[[i]], steiger_dir==TRUE)))
  loo[[i]] <- loores
}

# LOO plots are written to results directory

for (i in 1:length(loo)) {
  res1 <- loo[[i]]
  name <- paste0("LOO_",res1[[c(1, 1)]],"_",res1[[c(2, 1)]][1])
  p_save <- mr_leaveoneout_plot(as.data.frame(res1))[[1]]
  ggsave(plot = p_save,filename = paste0("../results/",name,".svg"))
}


# Single SNP analysis

single <- list()
for (i in 1:length(exposure_outcome)) {
  singleres <- data.frame(mr_singlesnp(subset(exposure_outcome[[i]], steiger_dir==TRUE)))
  single[[i]] <- singleres
}

# Funnel plots are written to results directory

for (i in 1:length(loo)) {
  res1 <- single[[i]]
  name <- paste0("Funnel_",res1[[c(1, 1)]],"_",res1[[c(2, 1)]])
  p_save <- mr_funnel_plot(as.data.frame(res1))[[1]]
  ggsave(p_save ,filename = paste0("../results/",name,".svg"))
}

# cochran's Q

Q <- list()
for (i in 1:length(exposure_outcome)) {
  Qres <- data.frame(mr_heterogeneity(subset(exposure_outcome[[i]], steiger_dir==TRUE)))
  Q[[i]] <- Qres
}

Q <- rbindlist(Q) 

```

As an example, here is the leave-one-out plot for healthspan:

![](../results/LOO_BMI_healthspan.svg){height=200%}   

And a funnel plot:

![](../results/Funnel_BMI_healthspan.svg)


Q-statistics: 

``` {r res Q}

kable(Q)

```

# Part 3: Different MR tools 

## MR-PRESSO

MR-PRESSO (Mendelian Randomization Pleiotropy RESidual Sum and Outlier) can be used to assess for and correct for horizontal pleiotropy. MR-PRESSO is computationally intense and should be run on a cluster. Unfortunately, decreasing the number of iterations often doesn't help (it needs more elements), so you can run it using parallel processing (requires furrr package).

```{r MR-PRESSO, eval=F}

presso <- function(x) {
   res <- mr_presso(data=data.frame(x),
                              BetaOutcome = "beta.outcome", 
                              BetaExposure = "beta.exposure", 
                              SdOutcome = "se.outcome", 
                              SdExposure = "se.exposure", 
                              OUTLIERtest = TRUE, DISTORTIONtest = TRUE,
                              NbDistribution = 10000, 
                              SignifThreshold = 0.05)
}

# Using the package "furrr" you can run this function in parallel using these commands:

# Get parallel program started. X is the number of processes, so the number of trait combinations you want to run in parallel
  plan(multisession, workers = 3)

# Run the analyses
  future_map(exposure_outcome, presso)


# Run it serially (does not require furrr):

# map(exposure_outcome, presso)

# Extract main result

presso_res <- list()
for (i in 1:length(res)) {
  presso_res[[i]] <- res[[i]]$`Main MR results`
}

# Give the elements the name of the outcome

names(presso_res) <- outcome_name_vector

presso_res <- rbindlist(presso_res, use.names=T, idcol = T) %>% 
  mutate(Exposure="MDD")

```



# Extra functions

If you want to automate your MR jobs, the functions below may be of help to run a larger number of analyses in the same go.

```{r OPTIONAL Automated instruments, eval=F}

############################### Optional

# Function for extracting hits
extract <- function(file, trait) {
  
  sumstats <- fread(file)
  
  sumstats <- sumstats %>% 
    mutate(exposure=trait) %>% 
    rename(effect_allele=A1, other_allele=A2, eaf=EAF, beta=B, se=SE, pval=P, samplesize=N)
  
  hits <- subset(sumstats, as.numeric(pval)<5e-08)
  
  fwrite(hits, file = paste0("sumstats/", trait, "_hits.txt"), sep=" ", na=NA, quote=F)
}

# Define file names
files <- c("xxx",
           "xxx",
           "xxx"
           )
# Name the traits
traits <- c("xxx", "xxx", "xxx")

# Run the function
map2(files, traits, extract)

# Read in the hit files that were just created 
files <- list.files(path="sumstats/", pattern=glob2rx("*_hits.txt"), full.name=T)

# Function for clumping the hit files
clump <- function(file) {
  x <- read_exposure_data(file,
                          phenotype_col = "exposure")
  y <- clump_data(x, clump_p1=5e-08)
  
  if (nrow(y)>10) # If there's fewer than 10 instruments when using a threshold of 5e-08, select SNPs with a p-value below 5e-7
    {res <- y}
  else
    {res <- clump_data(x, clump_p1=5e-7)}
  
  if (nrow(res)<10) # If there's still fewer than 10, further increase the threshold
  {res <- clump_data(x, clump_p1=5e-6)}
  
  if (nrow(res)<10)
  {res <- clump_data(x, clump_p1=5e-5)}

  tibble(res)
}

# Run the clumping function

exp_dat <- list()

for (i in 1:length(files)) {
  exp_dat[[i]] <- clump(files[[i]])
}

# Use the trait names to name the elements in the list
names(exp_dat) <- traits

# Count the number of instruments per exposure 
count <- list()
for (i in 1:length(exp_dat)) {
  count[[i]] <- exp_dat[[i]] %>% group_by(exposure) %>% 
    summarise(IV_SNPs=n())
}

```
