---
title: "MR tutorial II"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    collapsed: true
    theme: united
  author: "Joëlle Pasman, Gudmundur Einarsson"
  date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}

# Essential
library(data.table)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(MendelianRandomization)

# Working directory is root of 
knitr::opts_knit$set(root.dir = "../notebooks")
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, error=FALSE)

# Prep data function
prep_data <- function(exp_df, out_df) {
  beta_exposure <- exp_df$Beta
  pval_exposure <- exp_df$comb_Pval
  beta_outcome <- out_df$Beta
  pval_outcome <- out_df$comb_Pval
  annotations_rs <- exp_df$rsName
  
  x<-beta_exposure
  y<-beta_outcome
  sex<-abs(x/qnorm(pval_exposure/2, lower.tail = FALSE))
  sey<-abs(y/qnorm(pval_outcome/2, lower.tail = FALSE))
  
  data<-as.data.frame(cbind(x,y,sex,sey,
                            pval_exposure,
                            pval_outcome))
  data$annotations_rs <- annotations_rs
  
  for (i in 1:nrow(data)){
    if (data$x[i]<0){
      data$x[i]<-(-1)*data$x[i]
      data$y[i]<-(-1)*data$y[i]
    }
  }
  
  f<-subset(data,data$x!=0 & data$y!=0)
  return(f)
}

# Helper plot function
mr_results_plot <- function(data, mr_ivw_slope, mr_egger_slope, mr_egger_intercept,
                            title_text, subtitle_text, xlab_text, ylab_text, xlim_v) {
  g_p <- ggplot(data=data, aes(x=x, y=y)) +
    geom_pointrange(aes(ymin=y-1.96*sey, ymax=y+1.96*sey)) +
    geom_pointrange(aes(xmin=x-1.96*sex, xmax=x+1.96*sex)) +
    geom_abline(aes(colour="IVW",slope=mr_ivw_slope, intercept = 0),linetype="dotted",lwd=1.1) +
    geom_abline(aes(colour="MR-Egger",slope=mr_egger_slope, intercept = mr_egger_intercept),linetype="dotted",lwd=1.1) +
    scale_color_manual(name = "Regression Line", values = c("IVW" = "blue", "MR-Egger" = "red")) +
    theme_bw() +
    ggtitle(title_text, subtitle=subtitle_text)+
    theme(text = element_text(size = 14)) + xlab(xlab_text) + ylab(ylab_text) +
    geom_hline(yintercept = 0, linetype="dotted") + geom_vline(xintercept = 0, linetype="dotted") +
    theme(axis.line = element_line(colour = "black"), panel.border = element_blank()) +
    theme(legend.position = 'top') +
    geom_label_repel(aes(label = annotations),
                     size = 3.5,box.padding = unit(0.2, "lines"),point.padding = unit(0.2, "lines"))
  if(!is.null(xlim_v)) {
    g_p <- g_p + xlim(xlim_v)
  }
  return(g_p)
}

```

Here we simply load the data directly, i.e. it has already been harmonized. You can find the table in the supplementary material of the published manuscript (and here under `./tables`). Dorsalgia, the common backpain diagnosis has the code M54 and IDD has the code M51.

```{r Load data}
##############################################################
# Load instruments (already harmonized)
# M54 is Dorsalgia
# M51 is IDD
##############################################################

# Instruments and associations with exposure
dorsalgia_exp <- fread("../tables/M54_IS_DK_FIN_Exposure.txt")
IDD_exp <- fread("../tables/M51_IS_DK_FIN_Exposure.txt")

# Associations with outcome
dorsalgia_out_idd_exp <- fread("../tables/M54_UKB_Outcome_w_M51_as_exp.txt")
IDD_out_dorsalgia_exp <- fread("../tables/M51_UKB_Outcome_w_M54_as_exp.txt")

# Simple ordering 
data_dor_exp <- prep_data(exp_df = dorsalgia_exp,
                          out_df = IDD_out_dorsalgia_exp)
# Add some custom annotations
data_dor_exp$annotations <- ifelse(data_dor_exp$y < 0 | data_dor_exp$y> 0.07, 
                                   data_dor_exp$annotations_rs,"")

data_idd_exp <- prep_data(exp_df = IDD_exp,
                          out_df = dorsalgia_out_idd_exp)

data_idd_exp$annotations <- ifelse(data_idd_exp$y < 0 | data_idd_exp$x > 0.14, data_idd_exp$annotations_rs,"")

```

Here we demonstrate how the analysis can be performed in R, without any packages.

```{r Basic R approach}
##############################################################
# basic R approach
##############################################################

# Inverse variance weighted regression
mr_slope_1 <- lm(y ~ x-1, weights=1/sey^2, data = data_dor_exp)
mr_intercept_1 <- lm(y ~ 1, weights=1/sey^2, data = data_dor_exp)
mr_egger_1 <- lm(y ~ x, weights=1/sey^2, data = data_dor_exp)

gp <- mr_results_plot(data = data_dor_exp, 
                      mr_ivw_slope = mr_slope_1$coefficients[1], 
                      mr_egger_slope = mr_egger_1$coefficients[2],
                      mr_egger_intercept = mr_egger_1$coefficients[1],
                      title_text = "Dorsalgia variants", 
                      subtitle_text = paste0("Dorsalgia (IS,DK & FIN) Exposure and IDD (UK-Biobank) Outcome",
                                             "\nIVW estimate is : ", format(coef(mr_slope_1)[1], digits = 3)), 
                      xlab_text = "Exposure Dorsalgia-log(OR)", ylab_text = "Outcome IDD-log(OR)", xlim_v = c(0,0.1))
(gp)


###
# Reverse
###
mr_slope_2 <- lm(y ~ x-1, weights=1/sey^2, data = data_idd_exp)
mr_intercept_2 <- lm(y ~ 1, weights=1/sey^2, data = data_idd_exp)
mr_egger_2 <- lm(y ~ x, weights=1/sey^2, data = data_idd_exp)

gp2 <- mr_results_plot(data = data_idd_exp, 
                      mr_ivw_slope = mr_slope_2$coefficients[1], 
                      mr_egger_slope = mr_egger_2$coefficients[2],
                      mr_egger_intercept = mr_egger_2$coefficients[1],
                      title_text = "IDD variants", 
                      subtitle_text = paste0("IDD (IS,DK & FIN) Exposure and Dorsalgia (UK-Biobank) Outcome",
                                             "\nIVW estimate is : ", format(coef(mr_slope_2)[1], digits = 3)), 
                      xlab_text = "Exposure IDD-log(OR)", ylab_text = "Outcome Dorsalgia-log(OR)",
                      xlim_v = NULL)
(gp2)

# Combine in single plot
gp + gp2

```

Here we try out the `MendelianRandomization` package. We rerun the analysis and obtain the same point estimates.

```{r MendelianRandomization Package}
##############################################################
# MendelianRandomization package
##############################################################
dor_inp <- mr_input(bx = data_dor_exp$x,
                    by = data_dor_exp$y,
                    bxse = data_dor_exp$sex,
                    byse = data_dor_exp$sey)
ivw_res_1 <- mr_ivw(dor_inp)
egger_res_1 <- mr_egger(dor_inp)


gp <- mr_results_plot(data = data_dor_exp, 
                      mr_ivw_slope = ivw_res_1@Estimate, 
                      mr_egger_slope = egger_res_1@Estimate,
                      mr_egger_intercept = egger_res_1@Intercept,
                      title_text = "Dorsalgia variants", 
                      subtitle_text = paste0("Dorsalgia (IS,DK & FIN) Exposure and IDD (UK-Biobank) Outcome",
                                             "\nIVW estimate is : ", format(coef(mr_slope_1)[1], digits = 3)), 
                      xlab_text = "Exposure Dorsalgia-log(OR)", ylab_text = "Outcome IDD-log(OR)", xlim_v = c(0,0.1))
(gp)

idd_inp <- mr_input(bx = data_idd_exp$x,
                    by = data_idd_exp$y,
                    bxse = data_idd_exp$sex,
                    byse = data_idd_exp$sey)
ivw_res_2 <- mr_ivw(idd_inp)
egger_res_2 <- mr_egger(idd_inp)

gp2 <- mr_results_plot(data = data_idd_exp, 
                       mr_ivw_slope = ivw_res_2@Estimate, 
                       mr_egger_slope = egger_res_2@Estimate,
                       mr_egger_intercept = egger_res_2@Intercept,
                       title_text = "IDD variants", 
                       subtitle_text = paste0("IDD (IS,DK & FIN) Exposure and Dorsalgia (UK-Biobank) Outcome",
                                              "\nIVW estimate is : ", format(coef(mr_slope_2)[1], digits = 3)), 
                       xlab_text = "Exposure IDD-log(OR)", ylab_text = "Outcome Dorsalgia-log(OR)",
                       xlim_v = NULL)
(gp2)

# Combine in singleplot
gp + gp2

```

```{r Normal vs t-dist}
##############################################################
# t-test vs normal
# Let's observe the impact with the number of variants
##############################################################

idd_inp <- mr_input(bx = data_idd_exp$x,
                    by = data_idd_exp$y,
                    bxse = data_idd_exp$sex,
                    byse = data_idd_exp$sey)


ivw_res_norm <- mr_ivw(idd_inp)
ivw_res_t <- mr_ivw(idd_inp, distribution = "t-dist")

print("P-value with normal assumptions: ")
print(ivw_res_norm@Pvalue)

print("P-value with t-distribution: ")
print(ivw_res_t@Pvalue)

```

```{r Normal vs t-dist2}
##############################################################
# t-test vs normal
# Let's observe the impact with the number of variants
##############################################################

# Seed is number of first name letter in alphabet
my_first_letter <- "g"
seed_number <- which(letters %in% c(my_first_letter))
print(paste0("My seed number is: ", seed_number))
set.seed(seed_number)

# Sample random 20 instruments
subset_inds <- sample(1:nrow(data_idd_exp), size = 20)
idd_inp <- mr_input(bx = data_idd_exp$x[subset_inds],
                    by = data_idd_exp$y[subset_inds],
                    bxse = data_idd_exp$sex[subset_inds],
                    byse = data_idd_exp$sey[subset_inds])


ivw_res_norm <- mr_ivw(idd_inp)

print("Here is my estimate: ")
print(ivw_res_norm@Estimate)

# Write code for the rest of the numbers in the google drive document!

```